Finally Tagless Staged Interpreters for Simpler Typed Languages

### 2007-04-06 tagless/tagless-final.tex
lots more little edits everywhere

Our contributions are as follows.
\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles
   (\S\ref{variations}).
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state (\S\ref{state}).
\item We describe an approach to self\hyp interpretation compatible with the
  above (\S\ref{selfinterp}).  Self\hyp interpretation turned out harder than expected.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We further discuss related work in~\S\ref{related}.

@@

In contrast to \S\ref{intro}, we henceforth encode 
binding in object programs using higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices. This makes the encoding convenient and
also ensures that our object programs are closed.

@@

We embed this language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics|. The class is so named because its interface gives the syntax for
the object language and its instances give the semantics.
The parameter |repr| to the class has kind |* -> *|.
\begin{code}
class Symantics repr where

  int  :: Int  -> repr Int
  bool :: Bool -> repr Bool

@@

Comparing |Symantics| with Figure~\ref{fig:object}
shows how to represent every typed, closed object term in the
metalanguage. Moreover, the representation preserves types, in the following
sense.
\begin{prop}
If an object term has the object type~$t$, then its
representation in the metalanguage has the type 
|forall repr.| |Symantics repr => repr |$t$.
\end{prop}
Conversely, the type system of the metalanguage checks statically that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, such as if
we forget |app| somewhere above.

\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig

  type ('c, 'dv) repr

@@

  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
  val if_ : ('c, bool) repr -> (unit -> ('c, 'da) repr) -> (unit -> ('c, 'da) repr) -> ('c, 'da) repr

end
\end{code}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language}
\label{fig:ocaml-simple}
\end{figure*}

To embed the same object language in (Meta)OCaml, we replace the type
class |Symantics| and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.
Two differences between this signature and the |Symantics| class in
Haskell are: the additional type parameter |c|, an
environment classifier \citep{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| due to OCaml's being a call-by-value
language.

The functor below encodes our running examples |test1| and $\mathit{power}$.
\begin{code}
module EX(S: Symantics) = struct
  open S
  let test1 () =
    app (lam (fun x -> x)) (bool true)
  let testpowfix () = 
    lam (fun x -> fix (fun self -> lam (fun n ->
    if_ (leq n (int 0)) (fun () -> int 1)
        (fun () -> mul x
          (app self (add n (int (-1))))))))
  let testpowfix7 = 
     lam (fun x -> app (app (testpowfix ()) x)
                       (int 7))
end
\end{code}
The dummy argument to |test1| and |testpowfix| is an artefact of
MetaOCaml, again related to monomorphism: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
The value restriction then dictates that
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation.
\begin{code}
module R = struct
  (* absolutely no wrappers *)
  type ('c,'dv) repr = 'dv

@@

  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()

end
\end{code}

As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R|
module.
\begin{code}
module EXR = EX(R)
\end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\begin{prop}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{prop}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{prop}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{prop}
These propositions follow immediately from the soundness of the
metalanguage's type system.

@@

Inserting brackets and escapes appropriately into the
evaluator~|R| above yields the compiler~|C| below.
% this code does not have the 'sv parameter. It shows up later.
\begin{code}
module C = struct

  type ('c,'dv) repr = ('c,'dv) code

@@

  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f = 
    .<let rec self n = .~(f .<self>.) n in self>.

  let add e1 e2     = .<.~e1 + .~e2>.
  let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .<.~x <= .~y>.
  let if_ eb et ee = 
    .<if .~eb then .~(et ()) else .~(ee ())>.

end
\end{code}
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1|
\begin{code}
let module E = EX(C) in E.test1 ()
\end{code}
gives the code value |.<(fun x_6 -> x_6) true>.|
of inferred type |('c, bool) C.repr|.  Similarly, interpreting |testpowfix7|
with
\begin{code}
let module E = EX(C) in E.testpowfix7
\end{code}
gives the code value
\begin{code}
.<fun x_1 -> (fun x_2 ->
  let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1 
               else x_2 * (self_3 (x_5 + (-1))))
   n_4
  in self_3) x_1 7>.
\end{code}
with many apparent $\beta$-redexes.

@@

We have also implemented this compiler in Haskell. Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code given above. The
accompanying code gives the full details.

@@

Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of four
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension to the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

Roughly speaking, a partial evaluator interprets each object term either
to a static (present-stage) term, as produced by the evaluator~|R|, or
to a dynamic (future-stage) term, as produced by the compiler~|C|.  To
distinguish between static and dynamic terms, one might try
to define partial evaluation as the Haskell data type
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrInt :: P0 Int -> C Int
abstrInt (S0 r) = int (unR r)
abstrInt (E0 c) = c

abstrBool :: P0 Bool -> C Bool
abstrBool (S0 r) = bool (unR r)
abstrBool (E0 c) = c
\end{code}
We then start to define |P0| as an instance of the |Symantics| class.
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x)
  bool x = S0 (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add e1 e2 = E0 (add (abstrInt e1)
                      (abstrInt e2))
\end{code}
Integer and boolean literals are obviously immediate, present-stage
values. Addition yields a static term if and only if both operands
are static. If so, we use the evaluator~|R| to add the operands.
If not, we extract dynamic terms from the operands and add them
using the compiler~|C|.
Thus we use the overloaded functions |int| and |bool| from the |Symantics|
instances for |R| and~|C| to define those for~|P0|.

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool|) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\citep{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
f = app (lift f)}}

@@

\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig

  type ('c,'sv,'dv) repr

@@

\begin{figure}
\begin{floatrule}
\begin{code}
module P = struct

  type ('c,'sv,'dv) repr = {st: 'sv option;
                            dy: ('c,'dv) code}

  let abstr {dy = x} = x
  let pdyn x = {st = None; dy = x}

@@

  let if_ eb et ee = match eb with
  | {st = Some b} -> if b then et () else ee ()
  | _ -> pdyn (C.if_ (abstr eb) 
                     (fun () -> abstr (et ()))
                     (fun () -> abstr (ee ())))

  let lam f =
  {st = Some f; 
   dy = C.lam (fun x -> abstr (f (pdyn x)))}

@@

  let fix f = 
    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
    in let rec self = function
       | {st = Some _} as e -> app (f (lam self)) e
       | e -> pdyn (C.app fdyn (abstr e))
       in {st = Some self; dy = fdyn}

end
\end{code}
\end{floatrule}
\caption{Our partial evaluator (\texttt{mul} and \texttt{leq} are elided)}
\label{fig:pe}
\end{figure}

@@

@@

We are not the first to implement a typed interpreter for a typed
language -- \citet{laod93} use a typed version of the SK language
(and typeclasses) to implement a meta-circular interpreter (rather than
a self-interpreter).  Interestingly, their meta-circular interpreter
appears to be tagfree!  However, they could not have implemented a
compiler or a partial-evaluator in the same way since they rely quite
heavily on having injection-projection pairs (through reflection and
reification).

We have been unable to find a direct reference that establishes that
the \emph{typed} $\lambda$-calculus possesses a final coalgebra structure.
For the untyped $\lambda$-calculus, 
\citet{HonsellLenisa,honsell99coinductive} establish several interesting
results along this line.  Of particular interest is their use of
contexts with a hole \citep[p.13]{honsell99coinductive} for the definition
of \emph{observational equivalence}, which directly relates to our
use of terms with a hole as a context for our self-interpreter 
(\S\ref{selfinterp}).  The reader is directed to bibliography of
\citet{honsell99coinductive} for references to the foundational work
in this important area.  Particularly intriguing is the link to the
coinductive aspects of B\"{o}hm trees, as pointed out by
\citet{berarducci-models} and \citet[Example 4.3.4]{jacobs-coalgebra}.

>>>>

Our contributions are as follows.
\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles
   (\S\ref{variations}).
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state (\S\ref{state}).
\item We describe an approach to self\hyp interpretation compatible with the
  above (\S\ref{selfinterp}).  Self\hyp interpretation turned out harder than expected.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

In contrast to \S\ref{intro}, we encode binding using higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices. This makes the encoding convenient and
also ensures that our object programs are closed.

@@

We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|). The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\begin{code}
class Symantics repr where
  int  :: Int  -> repr Int
  bool :: Bool -> repr Bool

@@

Comparing |Symantics| with Figure~\ref{fig:object}
shows how to represent every typed, closed object term in the
metalanguage. Moreover, the representation preserves types, in the following
sense.
\begin{prop}
If an object term has the object type~$t$, then its
representation in the metalanguage has the type 
|forall repr.| |Symantics repr => repr |$t$.
\end{prop}
Conversely, the type system of the metalanguage statically checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, such as if
we forget |app| somewhere above.

\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig
  type ('c, 'dv) repr

@@

  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
  val if_ : ('c, bool) repr -> (unit -> ('c, 'da) repr) -> (unit -> ('c, 'da) repr) -> ('c, 'da) repr
end
\end{code}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language}
\label{fig:ocaml-simple}
\end{figure*}

To embed the same object language in (Meta)OCaml, we replace the type
class |Symantics| and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.  The two differences are:
the additional type parameter |c|, an
environment classifier \citep{WalidPOPL03} (required by MetaOCaml for
code generation in~\S\ref{compiler}); and the $\eta$-expanded type for
|fix| and thunk types in |if_| since OCaml is a call-by-value
language.

The functor below encodes our running examples |test1| and $\mathit{power}$.
\begin{code}
module EX(S: Symantics) = struct
  open S
  let test1 () =
    app (lam (fun x -> x)) (bool true)
  let testpowfix () = 
    lam (fun x -> fix (fun self -> lam (fun n ->
    if_ (leq n (int 0)) (fun () -> int 1)
        (fun () -> mul x
          (app self (add n (int (-1))))))))
  let testpowfix7 = 
     lam (fun x -> app (app (testpowfix ()) x)
                       (int 7))
end
\end{code}
The dummy argument to |test1| and |testpowfix| is an artefact of
MetaOCaml, related to monomorphism: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
The value restriction dictates that
the definitions of our object terms must look syntactically like
values. Alternatively, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation.
\begin{code}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)

@@

  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code}

As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|.
\begin{code}
module EXR = EX(R)
\end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\begin{prop}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{prop}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{prop}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{prop}
These propositions follow immediately from the soundness of the
metalanguage's type system.

@@

Inserting brackets and escapes appropriately into the
evaluator~|R| above yields the simple compiler~|C| below.
% this code does not have the 'sv parameter. It shows up later.
\begin{code}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code

@@

  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f = 
    .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2     = .<.~e1 + .~e2>.
  let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .<.~x <= .~y>.
  let if_ eb et ee = 
    .<if .~eb then .~(et ()) else .~(ee ())>.
end
\end{code}
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1|
\begin{code}
let module E = EX(C) in E.test1 ()
\end{code}
gives the code value |.<(fun x_6 -> x_6) true>.|
of inferred type |('c, bool) C.repr|.  Interpreting |testpowfix7|
with
\begin{code}
let module E = EX(C) in E.testpowfix7
\end{code}
gives a code value with many apparent $\beta$-redexes:
\begin{code}
.<fun x_1 -> (fun x_2 ->
  let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1 
               else x_2 * (self_3 (x_5 + (-1))))
   n_4 in self_3) x_1 7>.
\end{code}

@@

We have also implemented this compiler in Haskell. Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.

@@

Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of four
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

Roughly speaking, a partial evaluator interprets each object term as either
a static (present-stage) term (using the evaluator~|R|), or
a dynamic (future-stage) term, (using the compiler~|C|).  To
distinguish between static and dynamic terms, one might try
to define partial evaluation via the Haskell data type
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrInt :: P0 Int -> C Int
abstrInt (S0 r) = int (unR r)
abstrInt (E0 c) = c

abstrBool :: P0 Bool -> C Bool
abstrBool (S0 r) = bool (unR r)
abstrBool (E0 c) = c
\end{code}
We then start to define |P0| as an instance of the |Symantics| class.
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x)
  bool x = S0 (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add e1 e2 = E0 (add (abstrInt e1)
                      (abstrInt e2))
\end{code}
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (via~|R|) if and only if both operands
are static, otherwise we extract the dynamics terms from the operands and
add them using the compiler~|C|.  Uniformly, we use
the overloaded functions from the |Symantics|
instances for |R| and~|C| to define those for~|P0|.

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool| say) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\citep{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
f = app (lift f)}}

@@

\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr

@@

\begin{figure}
\begin{floatrule}
\begin{code}
module P = struct
  type ('c,'sv,'dv) repr = {st: 'sv option;
                            dy: ('c,'dv) code}
  let abstr {dy = x} = x
  let pdyn x = {st = None; dy = x}

@@

  let if_ eb et ee = match eb with
  | {st = Some b} -> if b then et () else ee ()
  | _ -> pdyn (C.if_ (abstr eb) 
                     (fun () -> abstr (et ()))
                     (fun () -> abstr (ee ())))
  let lam f =
  {st = Some f; 
   dy = C.lam (fun x -> abstr (f (pdyn x)))}

@@

  let fix f = 
    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
    in let rec self = function
       | {st = Some _} as e -> app (f (lam self)) e
       | e -> pdyn (C.app fdyn (abstr e))
       in {st = Some self; dy = fdyn}
end
\end{code}
\end{floatrule}
\caption{Our partial evaluator (\texttt{mul} and \texttt{leq} are elided)}
\label{fig:pe}
\end{figure}

@@

We are not the first to implement a typed interpreter for a typed
language -- \citet{laod93} use a typed version of the SK language
(and typeclasses) to implement a meta-circular interpreter (rather than
a self-interpreter).  Interestingly, their meta-circular interpreter
appears to be tagfree!  However, they could not have implemented a
compiler or a partial-evaluator in the same way since they rely quite
heavily on having injection-projection pairs.

@@

We have been unable to find work that establishes that
the \emph{typed} $\lambda$-calculus possesses a final coalgebra structure.
For the untyped $\lambda$-calculus, 
\citet{HonsellLenisa,honsell99coinductive} establish several interesting
results along this line.  Of particular interest is their use of
contexts with a hole \citep[p.13]{honsell99coinductive} for the definition
of \emph{observational equivalence}
(see \S\ref{selfinterp}).  The reader is directed to bibliography of
\citet{honsell99coinductive} for references to the foundational work
in this important area.  Particularly intriguing is the link to the
coinductive aspects of B\"{o}hm trees, as pointed out by
\citet{berarducci-models} and \citet[Example 4.3.4]{jacobs-coalgebra}.
